---
title: "Methylation_analysis_by_sex"
author: "Alos Diallo"
date: "2024-01-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}

library(minfi)
library(minfiData)
library(ENmix)
library(FlowSorted.Blood.EPIC)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(ggplot2)
library(ggpubr)
library(IlluminaHumanMethylationEPICmanifest)
library(dplyr)
library(limma)
library(ewastools)
library(data.table)
library(reshape2)  # For melting the data frame
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
require(grid)
library(qqman)
library(table1)
library(sesame)
library(ExperimentHub)
library("sesameData")
library(RPMM)
library(doParallel)
sesameDataCache()

```

```{r}
setwd("/Users/adiallo/Desktop/Dartmouth/Christiansen_lab/CF_Project/Methylation/combined")

# Location of idat files
idat <- "/Users/adiallo/Desktop/Dartmouth/Christiansen_lab/CF_Project/Methylation/combined"

```

By default it works for human, you have to set something else to have it work for mice
```{r}
betas = openSesame(idat, func = getBetas) 
```

Loading the annotation information 
```{r}
load("/Users/adiallo/Desktop/Dartmouth/Christiansen_lab/CF_Project/Methylation/Annotation/EPIC.hg19.manifest.RDATA")
load("/Users/adiallo/Desktop/Dartmouth/Christiansen_lab/CF_Project/Methylation/Annotation/FlowSorted.BloodExtended.EPIC.compTable.rda")
attach(annotation)


```

Sub-setting the probes
```{r}
use = annotation %>% filter(MASK_general == "FALSE" & NonCpG == "FALSE" & SexProbe == "FALSE")
use<-use$probeID
beta_subset<-betas[use,]
#mvals = minfi::getM(beta_subset)
```

Getting all of the files together 
```{r}
beta_files = mclapply(searchIDATprefixes("."),readIDATpair,mc.cores = 2)
```

Infering patient sex, age, and ethenicty 
```{r message=FALSE, warning=FALSE, include=FALSE}
targets = NULL
idat_file <- "/Users/adiallo/Desktop/Dartmouth/Christiansen_lab/CF_Project/Methylation/combined/sample_sheet"
targets <- read.metharray.sheet(idat_file)
genders = NULL


for(i in 1:length(beta_files)){
  genders[i] = inferSex(beta_files[[i]])
  
}
targets = cbind(targets,genders)

ethnicity = NULL

for(i in 1:length(beta_files)){
  ethnicity[i] = inferEthnicity(beta_files[[i]])
  
}


#infer age
ages = methyAge(beta_subset)
```


Attaching the different pieces of information to the dataframe.
```{r}
age = ages$mAge_Hovath
targets = cbind(targets,age)

colnames(ages)
targets = cbind(targets,ethnicity)



Age<-ages$mAge_Hovath
Sex<-factor(targets$genders)
Condition = factor(targets$Condition)
Cell_Type = factor(targets$Cell)
Patient = factor(targets$Patient)

```
```{r}
RGset <- read.metharray.exp(idat,force = TRUE)

MSet_noob <- preprocessNoob(RGset)
beta_matrix<-getBeta(MSet_noob)


#EPIC
Pred_EPIC <- projectCellType_CP(beta_matrix[rownames(FlowSorted.BloodExtended.EPIC.compTable),], 
                                FlowSorted.BloodExtended.EPIC.compTable,lessThanOne =T)*100
Pred_EPIC


#450K
Pred_450K <- projectCellType_CP(beta_matrix[rownames(IDOLOptimizedCpGs450klegacy.compTable),], 
                                IDOLOptimizedCpGs450klegacy.compTable,lessThanOne =T)*100



```


```{r}
#include the person in the model, so CF ID
#Think about adjusting for cell type also
design <- model.matrix(~0+Condition+Age+Sex)#adjusted for age,sex
colnames(design) <- c(levels(Condition),"Age",levels(Sex)[-1])

fit1 <- lmFit(beta_subset, design)


contMatrix1 <- makeContrasts(Post_Drug-Baseline,levels = design)

fit2 <- contrasts.fit(fit1, contMatrix1)
fit2 <- eBayes(fit2)


# look at the numbers of differentially methylated CpGs at FDR < 0.05
summary(decideTests(fit2,adjust.method = "fdr",p.value = 0.05))
```

Adjust for cell type:

```{r}
targets$Cell <- make.names(targets$Cell, unique=TRUE)
Cell_Type = factor(targets$Cell)
targets$CleanCell <- gsub("\\.\\d+", "", targets$Cell)
targets$CleanCell[targets$CleanCell == "T.cells"] <- "T.cell"
targets$CleanCell <- tolower(targets$CleanCell)
Cell_Type = factor(targets$CleanCell)
```

```{r}

design <- model.matrix(~ 0 + Condition + Age + Sex + Cell_Type)


colnames(design) <- c(levels(Condition), "Age", levels(Sex)[-1], levels(Cell_Type)[-1])
fit1 <- lmFit(beta_subset, design)
contMatrix1 <- makeContrasts(Post_Drug-Baseline, levels=design)
fit2 <- contrasts.fit(fit1, contMatrix1)
fit2 <- eBayes(fit2)
summary(decideTests(fit2, adjust.method="fdr", p.value=0.05))

```

```{r}
#Extract Results: Use the topTable function from the limma package to extract the top differentially methylated sites.
topResults <- topTable(fit2, coef="Post_Drug - Baseline", number=Inf)
# Filter by Adjusted p-value
sigResults <- topResults[topResults$adj.P.Val < 0.001,]

#To view the sites with the largest increases in methylation:
head(sigResults[order(-sigResults$logFC),])

#To view the sites with the largest decreases in methylation:
head(sigResults[order(sigResults$logFC),])

sigResults$probeID <- rownames(sigResults)
significantSites <- merge(sigResults, annotation, by="probeID", all.x=TRUE)

```

Setting up data 
```{r}
# Extracting the top CpGs using topTable
top_cpgs <- topTable(fit2, number=Inf, coef="Post_Drug - Baseline", sort.by="p", p.value=0.05)

# Filtering CpGs based on effect size and select top 1000
selected_cpgs <- top_cpgs[abs(top_cpgs$logFC) > 0.1, ][1:1000,]

# Subset betas
selected_beta <- beta_subset[rownames(beta_subset) %in% rownames(selected_cpgs), ]

selected_beta_no_na <- selected_beta[complete.cases(selected_beta), ]

# Assuming selected_beta_no_na has CpGs as rows and Samples as columns
long_data <- selected_beta_no_na %>% 
  as.data.frame() %>%
  rownames_to_column("CpG") %>%
  pivot_longer(cols = -CpG, names_to = "Sample", values_to = "Beta")


# Convert row names to a column in 'targets'
targets <- targets %>% mutate(Sample = row.names(targets))

# Perform the left join
long_data <- left_join(long_data, targets, by = c("Sample" = "Sample"))


```

```{r}

#Make the annotation dataframe for the heatmap
heat_annot <- data.frame(
  row.names= targets$Sample,
  Gender= targets$genders,
  Patient = targets$Patient,
  Treatment= targets$Condition,
  Tissue = targets$CleanCell
  )


# Define colors for annotation color bars
patient_colors <- c("AW" = "#CEEAB9", "GF" = "#8EF9D0", "GH" = "#5BACF7", "MC" = "#E9B1F7", "NG" = "#B48BE2")

ann_colors <- list(
  Gender = c('FEMALE' = "#F21F66", 'MALE' = "#176EBD"), 
  Patient = patient_colors,  # Use the patient_colors vector here
  Tissue = c('monocyte' = "#999BB2", 'neutrophil' = "#99B2A8", 'pbmc' = "#385CA2", 't.cell' = "#8A5851", 'whole.blood' = "#AE8B75"),
  Treatment = c('Baseline'= "grey", 'Post_Drug'= "black")
)

# Extracting the top CpGs using topTable
top_cpgs <- topTable(fit2, number=Inf, coef="Post_Drug - Baseline", sort.by="p", p.value=0.05)

# Filtering CpGs based on effect size and select top 1000
selected_cpgs <- top_cpgs[abs(top_cpgs$logFC) > 0.05, ][1:5000,]

# Subset betas
selected_beta <- beta_subset[rownames(beta_subset) %in% rownames(selected_cpgs), ]

selected_beta_no_na <- selected_beta[complete.cases(selected_beta), ]
# Calculate variable of each CpG across all samples(all subjects, all tissue types)
library(matrixStats)
CpG_Var <- matrixStats::rowVars(selected_beta_no_na)
rankVar <- data.frame('order' = rank(-CpG_Var), 'var'= CpG_Var)
rankVar$top10000 <- ifelse(rankVar$order <= 5000, 'Yes', 'No')
ggpubr::ggscatter(rankVar, x= 'order', y= 'var', color= 'top10000', palette= c('black', 'red') )

# Get rank of CpGs by
rankvar= rank(-CpG_Var)


# top 10,000 cpgs
data.topvar <- selected_beta_no_na[rankVar$top10000 == 'Yes',]
data.topvar <- as.matrix(data.topvar)


#6 blocks of 8 if plotting by chip there might not be enough cpg's 
pheatmap(
  data.topvar,
  annotation_names_col= T,
  show_rownames= FALSE, #CpGs
  show_colnames= TRUE, #Samples
  annotation_col= heat_annot, 
  #annotation_row= row_annot,
  annotation_colors= ann_colors,
  cluster_cols= TRUE,
  color= colorRampPalette(c("yellow", "blue"))(1024),
  clustering_distance_rows= "manhattan",
  clustering_distance_colors= "manhattan",
  clustering_method= "average",
  border_color= NA,
  fontsize= 13
)

```

Use this heatmap
```{r}
# Make the annotation dataframe for the heatmap with 'Patient' column
heat_annot <- data.frame(
  row.names = targets$FileName,
  Gender = targets$genders,
  Treatment= targets$Condition,
  Patient = targets$Patient,  # Changed Treatment to Patient
  Tissue = targets$CleanCel
)

# Order the heat_annot dataframe by Gender
heat_annot <- heat_annot[order(heat_annot$Gender), ]
# Define colors for annotation color bars
patient_colors <- c("AW" = "#CEEAB9", "GF" = "#8EF9D0", "GH" = "#5BACF7", "MC" = "#E9B1F7", "NG" = "#B48BE2")

ann_colors <- list(
  Gender = c('FEMALE' = "#F21F66", 'MALE' = "#176EBD"), 
  Treatment = c('Baseline'= "grey", 'Post_Drug'= "black"),
  Patient = patient_colors,  # Use the patient_colors vector here
  Tissue = c('monocyte' = "#999BB2", 'neutrophil' = "#99B2A8", 'pbmc' = "#385CA2", 't.cell' = "#8A5851", 'whole.blood' = "#AE8B75")
)

# Extracting the top CpGs using topTable
top_cpgs <- topTable(fit2, number=Inf, coef="Post_Drug - Baseline", sort.by="p", p.value=0.05)

# Filtering CpGs based on effect size and select top 1000
selected_cpgs <- top_cpgs[abs(top_cpgs$logFC) > 0.05, ][1:10000,]

# Subset betas
selected_beta <- beta_subset[rownames(beta_subset) %in% rownames(selected_cpgs), ]

selected_beta_no_na <- selected_beta[complete.cases(selected_beta), ]
# Calculate variable of each CpG across all samples(all subjects, all tissue types)
library(matrixStats)
CpG_Var <- matrixStats::rowVars(selected_beta_no_na)
rankVar <- data.frame('order' = rank(-CpG_Var), 'var'= CpG_Var)
rankVar$top10000 <- ifelse(rankVar$order <= 10000, 'Yes', 'No')
ggpubr::ggscatter(rankVar, x= 'order', y= 'var', color= 'top10000', palette= c('black', 'red') )

# Get rank of CpGs by
rankvar= rank(-CpG_Var)

#colnames(selected_beta_no_na) <- targets$Patient[match(colnames(selected_beta_no_na), targets$Sample)]

# top 10,000 cpgs
data.topvar <- selected_beta_no_na[rankVar$top10000 == 'Yes',]
data.topvar <- as.matrix(data.topvar)
#(selected_beta_no_na) <- targets$Patient[match(colnames(selected_beta_no_na), targets$Sample)]



# First, sort your targets dataframe by gender
# This will also sort the Sample column in the same order
targets_ordered_by_gender <- targets[order(targets$Condition, targets$genders), ]

# Create an ordered vector of sample names based on the gender order
ordered_sample_names <- targets_ordered_by_gender$FileName

# Now reorder the columns of your data matrix to match this order
# This assumes that your data matrix columns correspond to Sample names
ordered_data <- data.topvar[, ordered_sample_names]

#6 blocks of 8 if plotting by chip there might not be enough cpg's 
png("/Users/adiallo/Desktop/by_treatment_heatmap.png", width = 20, height = 20, units = "in", res = 600)
pheatmap(
  ordered_data,
  annotation_names_col= T,
  show_rownames= FALSE, #CpGs
  show_colnames= TRUE, #Samples
  annotation_col = heat_annot[ordered_sample_names, ],
  #annotation_row= row_annot,
  annotation_colors= ann_colors,
  cluster_cols= FALSE,
  color= colorRampPalette(c("yellow", "blue"))(1024),
  clustering_distance_rows= "manhattan",
  clustering_distance_colors= "manhattan",
  clustering_method= "average",
  border_color= NA,
  fontsize= 13
)
dev.off()
```


```{r}
# Order heat_annot correctly
heat_annot_ordered <- heat_annot[match(colnames(data.topvar), rownames(heat_annot)), ]

# Filter the data for baseline and post-drug treatments
baseline_data <- data.topvar[, heat_annot_ordered$Treatment == 'Baseline']
post_drug_data <- data.topvar[, heat_annot_ordered$Treatment == 'Post_Drug']

# Corresponding annotations for each subset
baseline_annot <- heat_annot_ordered[heat_annot_ordered$Treatment == 'Baseline', ]
post_drug_annot <- heat_annot_ordered[heat_annot_ordered$Treatment == 'Post_Drug', ]

# Check for NA, NaN, or Inf values and replace them if necessary
# For example, replacing NAs with the mean of non-NA values in the column
baseline_data[is.na(baseline_data)] <- apply(baseline_data, 2, function(x) mean(x, na.rm = TRUE))
post_drug_data[is.na(post_drug_data)] <- apply(post_drug_data, 2, function(x) mean(x, na.rm = TRUE))


# Check for NA, NaN, or Inf values in the baseline_data
sum(is.na(baseline_data))  # Should be 0 for no NAs
sum(is.nan(baseline_data)) # Should be 0 for no NaNs
sum(is.infinite(baseline_data)) # Should be 0 for no Infs


# Plot the baseline heatmap
#png("/Users/adiallo/Desktop/baseline_heatmap.png", width = 10, height = 10, units = "in", res = 300)
pheatmap(
  baseline_data,
  annotation_col = baseline_annot,
  annotation_colors = ann_colors,
  main = "Base Line",
  annotation_names_col= TRUE,
  show_rownames= FALSE,
  show_colnames= FALSE,
  cluster_cols= FALSE,  # You can cluster within the treatment group
  color = colorRampPalette(c("yellow", "blue"))(1024),
  border_color = NA,
  fontsize = 13
)
#dev.off()

# Plot the post-drug heatmap
#png("/Users/adiallo/Desktop/post_drug_heatmap.png", width = 10, height = 10, units = "in", res = 300)
pheatmap(
  post_drug_data,
  annotation_col = post_drug_annot,
  annotation_colors = ann_colors,
  main = "Post Drug",
  annotation_names_col= TRUE,
  show_rownames= FALSE,
  show_colnames= FALSE,
  cluster_cols= TRUE,  # You can cluster within the treatment group
  color = colorRampPalette(c("yellow", "blue"))(1024),
  border_color = NA,
  fontsize = 13
)
#dev.off()

```

By gender
```{r}
# Make the annotation dataframe for the heatmap with 'Patient' column
heat_annot <- data.frame(
  row.names = targets$Sample,
  Gender = targets$genders,
  Treatment= targets$Condition,
  Patient = targets$Patient,  # Changed Treatment to Patient
  Tissue = targets$CleanCel
)

# Order the heat_annot dataframe by Gender
heat_annot <- heat_annot[order(heat_annot$Gender), ]
# Define colors for annotation color bars
patient_colors <- c("AW" = "#CEEAB9", "GF" = "#8EF9D0", "GH" = "#5BACF7", "MC" = "#E9B1F7", "NG" = "#B48BE2")

ann_colors <- list(
  Gender = c('FEMALE' = "#F21F66", 'MALE' = "#176EBD"), 
  Treatment = c('Baseline'= "grey", 'Post_Drug'= "black"),
  Patient = patient_colors,  # Use the patient_colors vector here
  Tissue = c('monocyte' = "#999BB2", 'neutrophil' = "#99B2A8", 'pbmc' = "#385CA2", 't.cell' = "#8A5851", 'whole.blood' = "#AE8B75")
)

# Extracting the top CpGs using topTable
top_cpgs <- topTable(fit2, number=Inf, coef="Post_Drug - Baseline", sort.by="p", p.value=0.05)

# Filtering CpGs based on effect size and select top 10000
selected_cpgs <- top_cpgs[abs(top_cpgs$logFC) > 0.05, ][1:10000,]

# Subset betas
selected_beta <- beta_subset[rownames(beta_subset) %in% rownames(selected_cpgs), ]

selected_beta_no_na <- selected_beta[complete.cases(selected_beta), ]
# Calculate variable of each CpG across all samples(all subjects, all tissue types)
library(matrixStats)
CpG_Var <- matrixStats::rowVars(selected_beta_no_na)
rankVar <- data.frame('order' = rank(-CpG_Var), 'var'= CpG_Var)
rankVar$top10000 <- ifelse(rankVar$order <= 10000, 'Yes', 'No')
ggpubr::ggscatter(rankVar, x= 'order', y= 'var', color= 'top10000', palette= c('black', 'red') )

# Get rank of CpGs by
rankvar= rank(-CpG_Var)

#colnames(selected_beta_no_na) <- targets$Patient[match(colnames(selected_beta_no_na), targets$Sample)]

# top 10,000 cpgs
data.topvar <- selected_beta_no_na[rankVar$top10000 == 'Yes',]
data.topvar <- as.matrix(data.topvar)
#(selected_beta_no_na) <- targets$Patient[match(colnames(selected_beta_no_na), targets$Sample)]



# First, sort your targets dataframe by gender
# This will also sort the Sample column in the same order
targets_ordered_by_gender <- targets[order(targets$CleanCell, targets$Sample), ]

# Create an ordered vector of sample names based on the gender order
ordered_sample_names <- targets_ordered_by_gender$Sample

# Now reorder the columns of your data matrix to match this order
# This assumes that your data matrix columns correspond to Sample names
ordered_data <- data.topvar[, ordered_sample_names]

#6 blocks of 8 if plotting by chip there might not be enough cpg's 
#png("/Users/adiallo/Desktop/heatmap_gender.png", width = 10, height = 10, units = "in", res = 400)
pheatmap(
  ordered_data,
  annotation_names_col= T,
  show_rownames= FALSE, #CpGs
  show_colnames= FALSE, #Samples
  annotation_col = heat_annot[ordered_sample_names, ],
  #annotation_row= row_annot,
  annotation_colors= ann_colors,
  cluster_cols= FALSE,
  color= colorRampPalette(c("yellow", "blue"))(1024),
  clustering_distance_rows= "manhattan",
  clustering_distance_colors= "manhattan",
  clustering_method= "average",
  border_color= NA,
  fontsize= 13
)
#dev.off()
```



```{r}
# Sort the annotations within each treatment group by gender
baseline_annot_ordered_by_gender <- baseline_annot[order(baseline_annot$Gender), ]
post_drug_annot_ordered_by_gender <- post_drug_annot[order(post_drug_annot$Gender), ]

# Use the ordered annotations to reorder the columns of the data matrices
# The columns in the data matrices should be ordered according to the order in the annotations
baseline_data_ordered_by_gender <- baseline_data[, row.names(baseline_annot_ordered_by_gender)]
post_drug_data_ordered_by_gender <- post_drug_data[, row.names(post_drug_annot_ordered_by_gender)]


# Generate the heatmaps with data ordered by gender within each treatment group
png("/Users/adiallo/Desktop/baseline_heatmap_s.png", width = 10, height = 10, units = "in", res = 300)
pheatmap(
  baseline_data_ordered_by_gender,
  annotation_col = baseline_annot_ordered_by_gender,
  annotation_colors = ann_colors,
  main = "Base Line (Ordered by Gender)",
  annotation_names_col = TRUE,
  show_rownames = FALSE,
  show_colnames = TRUE,
  cluster_cols = FALSE,  # Disable clustering to maintain gender order
  color = colorRampPalette(c("yellow", "blue"))(1024),
  border_color = NA,
  fontsize = 13
)
dev.off()

png("/Users/adiallo/Desktop/post_drug_heatmap_s.png", width = 10, height = 10, units = "in", res = 300)
pheatmap(
  post_drug_data_ordered_by_gender,
  annotation_col = post_drug_annot_ordered_by_gender,
  annotation_colors = ann_colors,
  main = "Post Drug (Ordered by Gender)",
  annotation_names_col = TRUE,
  show_rownames = FALSE,
  show_colnames = TRUE,
  cluster_cols = FALSE,  # Disable clustering to maintain gender order
  color = colorRampPalette(c("yellow", "blue"))(1024),
  border_color = NA,
  fontsize = 13
)
dev.off()
```
Trying to make cell type heatmaps
```{r}
# First, sort your targets dataframe by gender
# This will also sort the Sample column in the same order
targets_ordered_by_condition <- targets[order(targets$Condition, targets$Sample), ]

# Create an ordered vector of sample names based on the gender order
ordered_sample_names <- targets_ordered_by_condition$Sample

# Now reorder the columns of your data matrix to match this order
# This assumes that your data matrix columns correspond to Sample names
# Now reorder the columns of your data matrix to match this order
# This assumes that your data matrix columns correspond to Sample names
ordered_data <- data.topvar[, ordered_sample_names]

# Order heat_annot correctly
heat_annot_ordered <- heat_annot[match(colnames(ordered_data), rownames(heat_annot)), ]

# Filter the data for baseline and post-drug treatments
monocyte_data <- ordered_data[, heat_annot_ordered$Tissue == 'monocyte']
neutrophil_data <- ordered_data[, heat_annot_ordered$Tissue == 'neutrophil']
tcell_data <- ordered_data[, heat_annot_ordered$Tissue == 't.cell']
whole.blood_data <- ordered_data[, heat_annot_ordered$Tissue == 'whole.blood']
pbmc_data <- ordered_data[, heat_annot_ordered$Tissue == 'pbmc']

# Corresponding annotations for each subset
monocyte_annot <- heat_annot_ordered[heat_annot_ordered$Tissue == 'monocyte', ]
neutrophil_annot <- heat_annot_ordered[heat_annot_ordered$Tissue == 'neutrophil', ]
pbmc_annot <- heat_annot_ordered[heat_annot_ordered$Tissue == 'pbmc', ]
wholeblood_annot <- heat_annot_ordered[heat_annot_ordered$Tissue == 'whole.blood', ]
tcell_annot <- heat_annot_ordered[heat_annot_ordered$Tissue == 't.cell', ]

# Check for NA, NaN, or Inf values and replace them if necessary
# For example, replacing NAs with the mean of non-NA values in the column
monocyte_data[is.na(monocyte_data)] <- apply(monocyte_data, 2, function(x) mean(x, na.rm = TRUE))
neutrophil_data[is.na(neutrophil_data)] <- apply(neutrophil_data, 2, function(x) mean(x, na.rm = TRUE))
tcell_data[is.na(tcell_data)] <- apply(tcell_data, 2, function(x) mean(x, na.rm = TRUE))
whole.blood_data[is.na(whole.blood_data)] <- apply(whole.blood_data, 2, function(x) mean(x, na.rm = TRUE))
pbmc_data[is.na(pbmc_data)] <- apply(pbmc_data, 2, function(x) mean(x, na.rm = TRUE))
# Check for NA, NaN, or Inf values in the baseline_data
sum(is.na(monocyte_data))  # Should be 0 for no NAs
sum(is.nan(monocyte_data)) # Should be 0 for no NaNs
sum(is.infinite(monocyte_data)) # Should be 0 for no Infs

sum(is.na(neutrophil_data))  # Should be 0 for no NAs
sum(is.nan(neutrophil_data)) # Should be 0 for no NaNs
sum(is.infinite(neutrophil_data)) # Should be 0 for no Infs

# Plot the baseline heatmap
png("/Users/adiallo/Desktop/monocyte_heatmap.png", width = 10, height = 10, units = "in", res = 300)
pheatmap(
  monocyte_data,
  annotation_col = monocyte_annot,
  annotation_colors = ann_colors,
  main = "Monocyte",
  annotation_names_col= TRUE,
  show_rownames= FALSE,
  show_colnames= TRUE,
  cluster_cols= TRUE,  # You can cluster within the treatment group
  color = colorRampPalette(c("yellow", "blue"))(1024),
  border_color = NA,
  fontsize = 13
)
dev.off()

# Plot the post-drug heatmap
png("/Users/adiallo/Desktop/neutrophil_heatmap.png", width = 10, height = 10, units = "in", res = 300)
pheatmap(
  neutrophil_data,
  annotation_col = neutrophil_annot,
  annotation_colors = ann_colors,
  main = "Neutrophil",
  annotation_names_col= TRUE,
  show_rownames= FALSE,
  show_colnames= TRUE,
  cluster_cols= TRUE,  # You can cluster within the treatment group
  color = colorRampPalette(c("yellow", "blue"))(1024),
  border_color = NA,
  fontsize = 13
)
dev.off()


# Plot the post-drug heatmap
png("/Users/adiallo/Desktop/tcell_heatmap.png", width = 10, height = 10, units = "in", res = 300)
pheatmap(
  tcell_data,
  annotation_col = tcell_annot,
  annotation_colors = ann_colors,
  main = "T-Cell",
  annotation_names_col= TRUE,
  show_rownames= FALSE,
  show_colnames= TRUE,
  cluster_cols= TRUE,  # You can cluster within the treatment group
  color = colorRampPalette(c("yellow", "blue"))(1024),
  border_color = NA,
  fontsize = 13
)
dev.off()

# Plot the post-drug heatmap
png("/Users/adiallo/Desktop/wb_heatmap.png", width = 10, height = 10, units = "in", res = 300)
pheatmap(
  whole.blood_data,
  annotation_col = wholeblood_annot,
  annotation_colors = ann_colors,
  main = "Whole Blood",
  annotation_names_col= TRUE,
  show_rownames= FALSE,
  show_colnames= TRUE,
  cluster_cols= TRUE,  # You can cluster within the treatment group
  color = colorRampPalette(c("yellow", "blue"))(1024),
  border_color = NA,
  fontsize = 13
)
dev.off()

# Plot the post-drug heatmap
png("/Users/adiallo/Desktop/pbmc_heatmap.png", width = 10, height = 10, units = "in", res = 300)
pheatmap(
  pbmc_data,
  annotation_col = pbmc_annot,
  annotation_colors = ann_colors,
  main = "PBMC",
  annotation_names_col= TRUE,
  show_rownames= FALSE,
  show_colnames= TRUE,
  cluster_cols= TRUE,  # You can cluster within the treatment group
  color = colorRampPalette(c("yellow", "blue"))(1024),
  border_color = NA,
  fontsize = 13
)
dev.off()

getwd()
```



Trying to now use RPMM clustering
```{r}

sample_names <- colnames(beta_subset)
getRPMMClustLabels <- function(rpmmObject, Y_inv=NULL) {
  #'@description Extracts RPMM hard cluster labels
  #'@param rpmmObject RPMM object
  #'@param Y_inv Optional. Input matrix for RPMM computation. If provided, the sample name will be updated
  
  hardLabels <- blcTreeLeafClasses(rpmmObject);
  hardLabels <- as.data.frame(hardLabels);
  colnames(hardLabels) <- "RPMMClusters";
  if(! is.null(Y_inv)) rownames(hardLabels) <- hardLabels$Sample_Name <- rownames(Y_inv);
  return(hardLabels);
}

getRPMMSampOrder <- function(rpmmClusters, Y_inv) {
  #'@description Retrieves sample orders fo heat map visualization
  #'@describeIn Hinoue et al. online tutorial
  #'@param rpmmClusters data.frame with row.names = sample ID & 1 column named RPMM with cluster assignments
  #'@param Y_inv Input matrix for RPMM computation
  
  sampOrder <- c();
  for(r in names(table(rpmmClusters$RPMM))) {
    samps <- rownames(rpmmClusters)[rpmmClusters$RPMM == r];
    clu <- t(Y_inv[samps, ]);
    s_i <- seriation::seriate(clu, margin = 2)
    so_i <- seriation::get_order(s_i, dim = 2)
    sampOrder <- c(sampOrder, samps[so_i]);
  }
  sampOrder <- data.frame(
    Sample_Name = sampOrder,
    RPMMSampleOrder = 1:length(sampOrder)
  );
  return(sampOrder);
}

samples_data = t((selected_beta_no_na))
sup_rpmm <- blcTree(samples_data, verbose=1, maxlevel=2); 
  print(sup_rpmm);

  ## Extract RPMM cluster labels:
  rpmmClusters <- getRPMMClustLabels(sup_rpmm, samples_data);
  
  ## Retrieve RPMM sample order:
  sampOrders <- getRPMMSampOrder(rpmmClusters, samples_data);  

  ## Export:
  dat_sup <- merge(rpmmClusters, sampOrders, by="Sample_Name");
  write.csv(dat_sup, file= "RPMM_clustering_results.csv")

RPMM <- dat_sup[,-1]

targets <- merge(targets, dat_sup, by.x= "Sample", by.y= "Sample_Name")

# Make the annotation dataframe for the heatmap with 'Patient' column
heat_annot <- data.frame(
  `RPMM Clusters` = targets$RPMMClusters,
  Tissue = targets$CleanCel,
  row.names = targets$Sample,
  Gender = targets$genders,
  Treatment= targets$Condition,
  Patient = targets$Patient  # Changed Treatment to Patient
)

# Define colors for annotation color bars
patient_colors <- c("AW" = "#CEEAB9", "GF" = "#8EF9D0", "GH" = "#5BACF7", "MC" = "#E9B1F7", "NG" = "#B48BE2")

ann_colors <- list(
  Gender = c('FEMALE' = "#F21F66", 'MALE' = "#176EBD"), 
  Treatment = c('Baseline'= "grey", 'Post_Drug'= "black"),
  Patient = patient_colors,  # Use the patient_colors vector here
  Tissue = c('monocyte' = "#999BB2", 'neutrophil' = "#99B2A8", 'pbmc' = "#385CA2", 't.cell' = "#8A5851", 'whole.blood' = "#AE8B75"),
  `RPMM Clusters` = c('rLL' = '#506C58', 'rLR' = '#76EBF2', 'rRL' = '#58085B', 'rRR' = '#FFA791')
)

# Create an ordered vector of sample names based on the RPMM cluster order
ordered_sample_names <- targets$Sample[order(targets$CleanCell)]

# Ensure that your data matrix has samples as columns and CpGs as rows
# If it's not already like that, you might need to transpose it
# data.topvar <- t(data.topvar)

# Reorder the columns of your data matrix to match the order of the samples
ordered_data <- data.topvar[, ordered_sample_names]

# Generate the heatmap with ordered data
# Define the size of the plot
width <- 10 # width in inches
height <- 10 # height in inches
filename <- "/Users/adiallo/Desktop/Clustering_Heatmap.png"

# Open a file to plot to
png(filename, width = width, height = height, units = "in", res = 300)

# Assuming data_matrix is your methylation data matrix and you want to set new column names
#colnames(ordered_data) <- targets$SubjectID
# Generate the heatmap

pheatmap(
  ordered_data,
  annotation_col = heat_annot[ordered_sample_names, ],
  annotation_colors = ann_colors,
  color = colorRampPalette(c("yellow", "blue"))(1024),
  border_color = NA,
  fontsize = 10,
  show_rownames = FALSE,
  show_colnames = TRUE,
  cluster_cols = FALSE
)

# Close the file
dev.off()


```

Heatmaps for cell types:
```{r}
# Make the annotation dataframe for the heatmap with 'Patient' column
heat_annot <- data.frame(
  `RPMM Clusters` = targets$RPMMClusters,
  Tissue = targets$CleanCel,
  row.names = targets$Sample,
  Gender = targets$genders,
  Treatment= targets$Condition,
  Patient = targets$Patient  # Changed Treatment to Patient
)

# Define colors for annotation color bars
patient_colors <- c("AW" = "#CEEAB9", "GF" = "#8EF9D0", "GH" = "#5BACF7", "MC" = "#E9B1F7", "NG" = "#B48BE2")

ann_colors <- list(
  Gender = c('FEMALE' = "#F21F66", 'MALE' = "#176EBD"), 
  Treatment = c('Baseline'= "grey", 'Post_Drug'= "black"),
  Patient = patient_colors,  # Use the patient_colors vector here
  Tissue = c('monocyte' = "#999BB2", 'neutrophil' = "#99B2A8", 'pbmc' = "#385CA2", 't.cell' = "#8A5851", 'whole.blood' = "#AE8B75"),
  `RPMM Clusters` = c('rLL' = '#506C58', 'rLR' = '#76EBF2', 'rRL' = '#58085B', 'rRR' = '#FFA791')
)

# Create an ordered vector of sample names based on the RPMM cluster order
ordered_sample_names <- targets$Sample[order(targets$CleanCell)]
# Function to generate heatmap for a given tissue
generate_heatmap_for_tissue <- function(tissue_name, data, annotation, ann_colors) {
  tissue_data <- data[annotation$Tissue == tissue_name, ]
  tissue_annotation <- annotation[annotation$Tissue == tissue_name, ]
  
  heatmap_filename <- paste0("/Users/adiallo/Desktop/heatmap_", tissue_name, ".png")
  png(heatmap_filename, width = 800, height = 600)  # Adjust size as needed
  
  pheatmap(
    tissue_data,
    annotation_col = tissue_annotation,
    annotation_colors = ann_colors,
    cluster_cols = TRUE,
    show_rownames = FALSE,
    show_colnames = FALSE,
    color = colorRampPalette(c("yellow", "blue"))(1024),
    clustering_distance_rows = "manhattan",
    clustering_distance_cols = "manhattan",
    clustering_method = "average",
    border_color = NA,
    fontsize = 13
  )
  
  dev.off() # Close the device
}

# Now, you can call this function for each tissue type
tissues <- c('monocyte', 'neutrophil', 'pbmc', 't.cell', 'whole.blood')

for(tissue in tissues) {
  generate_heatmap_for_tissue(tissue, data.topvar, heat_annot, ann_colors)
}
```


Generating plots of celltype proportions 
```{r}
# Convert your matrix to a data frame in long format
df <- as.data.frame(Pred_EPIC)
df$Sample <- rownames(df)
long_df <- tidyr::gather(df, CellType, Proportion, -Sample)

# Add patient information
long_df$Patient <- targets$Patient[match(long_df$Sample, rownames(Pred_EPIC))]
long_df$genders <- targets$genders[match(long_df$Sample, rownames(Pred_EPIC))]

# Create a boxplot
p <- ggplot(long_df, aes(x = CellType, y = Proportion, color = genders)) +
  geom_boxplot(outlier.shape = NA) + # do not show outliers twice
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9), alpha = 0.5) +
  theme_minimal() +
  labs(title = "Cell Type Proportions by Sex", y = "Proportion (%)", x = "Cell Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

options(repr.plot.width=20, repr.plot.height=20)
print(p)
```

Gnerating average plots
```{r}
# Create the plot with points
p <- ggplot(long_df, aes(x=CellType, y=Proportion)) +
  geom_point(position = position_jitter(width = 0.2), aes(color=CellType), size=1) +  # jittered points
  theme_minimal() +
  labs(title="Scatter Plot of Cell Types") +
  facet_wrap(~genders)  # separate plot for each patient

# Display the plot with custom dimensions
options(repr.plot.width=20, repr.plot.height=6)
print(p)
```
```{r}
# Create the plot with points, stacking the plots vertically
p <- ggplot(long_df, aes(x=CellType, y=Proportion, color=CellType)) +
  geom_point(position = position_jitter(width = 0.2), size=1) +  # jittered points
  theme_minimal() +
  labs(title="Scatter Plot of Cell Types") +
  facet_wrap(~genders, ncol = 1) +  # Stack plots vertically
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),  # Rotate x-axis text
    strip.text = element_text(size = 12)  # Adjust facet label text size if needed
  )

# Adjust the dimensions of the plot to fit the stacked layout
options(repr.plot.width=12, repr.plot.height=12)
print(p)

```


```{r}
# Convert your matrix to a data frame in long format
df <- as.data.frame(Pred_EPIC)
df$Sample <- rownames(df)
long_df <- tidyr::gather(df, CellType, Proportion, -Sample)

# Add patient information
long_df$Patient <- targets$Patient[match(long_df$Sample, rownames(Pred_EPIC))]
long_df$genders <- targets$genders[match(long_df$Sample, rownames(Pred_EPIC))]

# Average the Proportion for each Patient and CellType combination
library(dplyr)
long_df_avg <- long_df %>%
  group_by(genders, CellType) %>%
  summarise(Proportion = mean(Proportion, na.rm = TRUE)) %>%
  ungroup()

# Create a plot using only points for the averaged data
p <- ggplot(long_df_avg, aes(x = CellType, y = Proportion, color = genders)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9), alpha = 0.5) +
  theme_minimal() +
  labs(title = "Averaged Cell Type Proportions by Sex", y = "Proportion (%)", x = "Cell Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

options(repr.plot.width=20, repr.plot.height=20)
print(p)


```
Showing both baseline and post-drug
```{r}

# Add condition information to the reshaped dataframe
long_df$Condition <- targets$Condition[match(long_df$Sample, rownames(Pred_EPIC))]
long_df$genders <- targets$genders[match(long_df$Sample, rownames(Pred_EPIC))]

# Group by Patient, CellType, and Condition to compute average proportions
long_df_avg <- long_df %>%
  group_by(genders, CellType, Condition) %>%
  summarise(Proportion = mean(Proportion, na.rm = TRUE)) %>%
  ungroup()

# Plot the data using points and distinguish between pre and post drug conditions using different shapes
p <- ggplot(long_df_avg, aes(x = CellType, y = Proportion, color = genders, shape = Condition)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9), alpha = 0.5) +
  theme_minimal() +
  labs(title = "Averaged Cell Type Proportions by Sex and Condition", y = "Proportion (%)", x = "Cell Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

options(repr.plot.width=20, repr.plot.height=20)
print(p)
```

```{r}


# Reshape the long_df_avg to get both Baseline and Post_Drug proportions on the same row
diff_df <- long_df_avg %>%
  spread(key = Condition, value = Proportion) %>%
  mutate(Difference = `Post_Drug` - `Baseline`) %>%
  filter(!is.na(Difference))  # remove rows without a computed difference

# Plot the differences
p <- ggplot(diff_df, aes(x = CellType, y = Difference, color = Patient)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9), alpha = 0.5) +
  theme_minimal() +
  labs(title = "Difference in Cell Type Proportions (Post Drug - Baseline)", y = "Proportion Difference (%)", x = "Cell Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#options(repr.plot.width=20, repr.plot.height=20)
print(p)


```


```{r}
# List of cell types
cell_types <- levels(factor(long_df$CellType))

# Generate a list of plots
plot_list <- list()

for (cell in cell_types) {
  # Filter data for the specific cell type
  filtered_data <- long_df[long_df$CellType == cell,]
  
  # Create the plot for the specific cell type
  p <- ggplot(filtered_data, aes(x = Patient, y = Proportion, fill = Patient)) +
    geom_boxplot() +
    labs(title = paste("Distribution of", cell), y = "Proportion (%)", x = "Patient") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  # Add the plot to the list
  plot_list[[cell]] <- p
}

# Print each plot
for (cell in cell_types) {
  print(plot_list[[cell]])
}
p <- ggplot(long_df, aes(x = genders, y = Proportion, fill = genders)) +
  geom_boxplot() +
  labs(title = "Distribution of Cell Types", y = "Proportion (%)", x = "Sex") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~ CellType, scales = "free_y", ncol = 4)  # Use facet_wrap to create a facet for each cell type

print(p)


```


```{r}
plot_list_diff = NULL
# Reshape the long_df_avg to get both Baseline and Post_Drug proportions on the same row
diff_df <- long_df_avg %>%
  spread(key = Condition, value = Proportion) %>%
  mutate(Difference = `Post_Drug` - `Baseline`) %>%
  filter(!is.na(Difference))  # remove rows without a computed difference

# Adjusting the plotting loop to work with diff_df
for (cell in cell_types) {
  # Filter data for the specific cell type
  filtered_data <- diff_df[diff_df$CellType == cell,]
  
  # Create the plot for the specific cell type
  p <- ggplot(filtered_data, aes(x = Patient, y = Difference, fill = Patient)) +
    geom_boxplot() +
    labs(title = paste("Difference in", cell, "(Post Drug - Baseline)"), y = "Proportion Difference (%)", x = "Patient") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  # Add the plot to the list
  plot_list_diff[[cell]] <- p
}

# Print each plot
for (cell in cell_types) {
  print(plot_list_diff[[cell]])
}

p <- ggplot(diff_df, aes(x = Patient, y = Difference, color = Patient)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9), alpha = 0.5) +
  labs(title = "Difference in Cell Types (Post Drug - Baseline)", y = "Proportion Difference (%)", x = "Patient") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~ CellType, scales = "free_y", ncol = 4)  # Use facet_wrap to create a facet for each cell type

print(p)



```




```{r}
# Merge to get TimePoint information into long_df
merged_data <- merge(long_df, targets[, c("Sample", "TimePoint")], by = "Sample")

# Assuming merged_data is correctly merged with TimePoint and other necessary information
# Filter for the 'Mono' cell type across all conditions, if you want to see the change over time regardless of condition
specific_cell_data <- merged_data %>%
  filter(CellType == "Eos")

# Create the boxplot for monocyte proportions by TimePoint with customized axis labels
p <- ggplot(specific_cell_data, aes(x = as.factor(TimePoint), y = Proportion, fill = as.factor(TimePoint))) +
  geom_boxplot() +
  labs(title = "Eosinophil Proportions by Timepoint", y = "Proportion (%)", x = "Time Point") +
  scale_x_discrete(labels = c("0" = "Base Line", "1" = "Post Drug Time Point 1", "2" = "Post Drug Time Point 2")) +
  theme_minimal()

print(p)

```

