---
title: "R Notebook"
output: html_notebook
---


```{r message=FALSE, warning=FALSE}

library(minfi)
library(minfiData)
library(ENmix)
library(FlowSorted.Blood.EPIC)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(ggplot2)
library(ggpubr)
library(IlluminaHumanMethylationEPICmanifest)
library(dplyr)
library(limma)
library(ewastools)
library(data.table)
library(reshape2)  # For melting the data frame
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
require(grid)
library(qqman)
library(table1)
library(sesame)
library(ExperimentHub)
library("sesameData")
library(RPMM)
library(doParallel)
sesameDataCache()

```

```{r}
setwd("/Users/adiallo/Desktop/Dartmouth/Christiansen_lab/CF_Project/Methylation/combined")

# Location of idat files
idat <- "/Users/adiallo/Desktop/Dartmouth/Christiansen_lab/CF_Project/Methylation/combined"

```

By default it works for human, you have to set something else to have it work for mice
```{r}
betas = openSesame(idat, func = getBetas) 
```

Loading the annotation information 
```{r message=FALSE, warning=FALSE}
load("/Users/adiallo/Desktop/Dartmouth/Christiansen_lab/CF_Project/Methylation/Annotation/EPIC.hg19.manifest.RDATA")
load("/Users/adiallo/Desktop/Dartmouth/Christiansen_lab/CF_Project/Methylation/Annotation/FlowSorted.BloodExtended.EPIC.compTable.rda")
attach(annotation)


```

Subsetting the probes -- Go back and look at the use var and make sure that we drop the sex data
```{r}
use = annotation %>% filter(MASK_general == "FALSE" & NonCpG == "FALSE" & SexProbe == "FALSE")
use<-use$probeID
beta_subset<-betas[use,]
#mvals = minfi::getM(beta_subset)
```

Getting all of the files together 
```{r}
beta_files = mclapply(searchIDATprefixes("."),readIDATpair,mc.cores = 2)
```

Infering patient sex, age, and ethenicty 
```{r message=FALSE, warning=FALSE, include=FALSE}
targets = NULL
idat_file <- "/Users/adiallo/Desktop/Dartmouth/Christiansen_lab/CF_Project/Methylation/combined/sample_sheet"
targets <- read.metharray.sheet(idat_file)
genders = NULL


for(i in 1:length(beta_files)){
  genders[i] = inferSex(beta_files[[i]])
  
}
targets = cbind(targets,genders)

ethnicity = NULL

for(i in 1:length(beta_files)){
  ethnicity[i] = inferEthnicity(beta_files[[i]])
  
}


#infer age
ages = methyAge(beta_subset)
```


Attaching the different pieces of information to the dataframe.
```{r}
age = ages$mAge_Hovath
targets = cbind(targets,age)

colnames(ages)
targets = cbind(targets,ethnicity)



Age<-ages$mAge_Hovath
Sex<-factor(targets$genders)
Condition = factor(targets$Condition)
Cell_Type = factor(targets$Cell)
Patient = factor(targets$Patient)

```
### Make sure to drop sex chromasomes if you have not, when you do the filtering.


12 Cell types:
```{r}
RGset <- read.metharray.exp(idat,force = TRUE)

MSet_noob <- preprocessNoob(RGset)
beta_matrix<-getBeta(MSet_noob)


#EPIC
Pred_EPIC <- projectCellType_CP(beta_matrix[rownames(FlowSorted.BloodExtended.EPIC.compTable),], 
                                FlowSorted.BloodExtended.EPIC.compTable,lessThanOne =T)*100
Pred_EPIC

write.table(Pred_EPIC,file="/Users/adiallo/Desktop/Pred_EPIC.txt",sep = "\t")
#450K
Pred_450K <- projectCellType_CP(beta_matrix[rownames(IDOLOptimizedCpGs450klegacy.compTable),], 
                                IDOLOptimizedCpGs450klegacy.compTable,lessThanOne =T)*100

write.table(Pred_450K,file="/Users/adiallo/Desktop/Pred_450k.txt",sep = "\t")

write.csv(targets, "/Users/adiallo/Desktop/Sample_sheet_infered_data.csv", row.names=FALSE)


install.packages("ggplot2")
install.packages("reshape2")
library(ggplot2)
library(reshape2)

# Reshape data into long format
long_data <- melt(Pred_EPIC)

# Generate heatmap
ggplot(long_data, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Sample", y = "Cell Type", fill = "Percentage")

```


```{r}
#include the person in the model, so CF ID
#Think about adjusting for cell type also
design <- model.matrix(~0+Condition+Age+Sex)#adjusted for age,sex
colnames(design) <- c(levels(Condition),"Age",levels(Sex)[-1])

fit1 <- lmFit(beta_subset, design)


contMatrix1 <- makeContrasts(Post_Drug-Baseline,levels = design)

fit2 <- contrasts.fit(fit1, contMatrix1)
fit2 <- eBayes(fit2)


# look at the numbers of differentially methylated CpGs at FDR < 0.05
summary(decideTests(fit2,adjust.method = "fdr",p.value = 0.05))
```

Adjust for cell type:

```{r}
targets$Cell <- make.names(targets$Cell, unique=TRUE)
Cell_Type = factor(targets$Cell)
targets$CleanCell <- gsub("\\.\\d+", "", targets$Cell)
targets$CleanCell[targets$CleanCell == "T.cells"] <- "T.cell"
targets$CleanCell <- tolower(targets$CleanCell)
Cell_Type = factor(targets$CleanCell)
```

```{r}

design <- model.matrix(~ 0 + Condition + Age + Sex + Cell_Type)


colnames(design) <- c(levels(Condition), "Age", levels(Sex)[-1], levels(Cell_Type)[-1])
fit1 <- lmFit(beta_subset, design)
contMatrix1 <- makeContrasts(Post_Drug-Baseline, levels=design)
fit2 <- contrasts.fit(fit1, contMatrix1)
fit2 <- eBayes(fit2)
summary(decideTests(fit2, adjust.method="fdr", p.value=0.05))

```

```{r}
#Extract Results: Use the topTable function from the limma package to extract the top differentially methylated sites.
topResults <- topTable(fit2, coef="Post_Drug - Baseline", number=Inf)
# Filter by Adjusted p-value
sigResults <- topResults[topResults$adj.P.Val < 0.001,]

#To view the sites with the largest increases in methylation:
head(sigResults[order(-sigResults$logFC),])

#To view the sites with the largest decreases in methylation:
head(sigResults[order(sigResults$logFC),])

sigResults$probeID <- rownames(sigResults)
significantSites <- merge(sigResults, annotation, by="probeID", all.x=TRUE)

```

Setting up data 
```{r}
# Extracting the top CpGs using topTable
top_cpgs <- topTable(fit2, number=Inf, coef="Post_Drug - Baseline", sort.by="p", p.value=0.05)

# Filtering CpGs based on effect size and select top 1000
selected_cpgs <- top_cpgs[abs(top_cpgs$logFC) > 0.1, ][1:1000,]

# Subset betas
selected_beta <- beta_subset[rownames(beta_subset) %in% rownames(selected_cpgs), ]

selected_beta_no_na <- selected_beta[complete.cases(selected_beta), ]

# Assuming selected_beta_no_na has CpGs as rows and Samples as columns
long_data <- selected_beta_no_na %>% 
  as.data.frame() %>%
  rownames_to_column("CpG") %>%
  pivot_longer(cols = -CpG, names_to = "Sample", values_to = "Beta")



# Perform the left join
#long_data <- left_join(long_data, targets, by = "Sample")

```

```{r}

#Make the annotation dataframe for the heatmap
heat_annot <- data.frame(
  row.names= targets$FileName,
  Gender= targets$genders,
  Patient = targets$Patient,
  Treatment= targets$Condition,
  Tissue = targets$CleanCell
  )


# Define colors for annotation color bars
patient_colors <- c("AW" = "#CEEAB9", "GF" = "#8EF9D0", "GH" = "#5BACF7", "MC" = "#E9B1F7", "NG" = "#B48BE2")

ann_colors <- list(
  Gender = c('FEMALE' = "#F21F66", 'MALE' = "#176EBD"), 
  Patient = patient_colors,  # Use the patient_colors vector here
  Tissue = c('monocyte' = "#999BB2", 'neutrophil' = "#99B2A8", 'pbmc' = "#385CA2", 't.cell' = "#8A5851", 'whole.blood' = "#AE8B75"),
  Treatment = c('Baseline'= "grey", 'Post_Drug'= "black")
)

# Extracting the top CpGs using topTable
top_cpgs <- topTable(fit2, number=Inf, coef="Post_Drug - Baseline", sort.by="p", p.value=0.05)

# Filtering CpGs based on effect size and select top 1000
selected_cpgs <- top_cpgs[abs(top_cpgs$logFC) > 0.05, ][1:1000,]

# Subset betas
selected_beta <- beta_subset[rownames(beta_subset) %in% rownames(selected_cpgs), ]

selected_beta_no_na <- selected_beta[complete.cases(selected_beta), ]
# Calculate variable of each CpG across all samples(all subjects, all tissue types)
library(matrixStats)
CpG_Var <- matrixStats::rowVars(selected_beta_no_na)
rankVar <- data.frame('order' = rank(-CpG_Var), 'var'= CpG_Var)
rankVar$top10000 <- ifelse(rankVar$order <= 1000, 'Yes', 'No')
ggpubr::ggscatter(rankVar, x= 'order', y= 'var', color= 'top10000', palette= c('black', 'red') )

# Get rank of CpGs by
rankvar= rank(-CpG_Var)


# top 10,000 cpgs
data.topvar <- selected_beta_no_na[rankVar$top10000 == 'Yes',]
data.topvar <- as.matrix(data.topvar)


#6 blocks of 8 if plotting by chip there might not be enough cpg's 
pheatmap(
  data.topvar,
  annotation_names_col= T,
  show_rownames= FALSE, #CpGs
  show_colnames= TRUE, #Samples
  annotation_col= heat_annot, 
  #annotation_row= row_annot,
  annotation_colors= ann_colors,
  cluster_cols= TRUE,
  color= colorRampPalette(c("yellow", "blue"))(1024),
  clustering_distance_rows= "manhattan",
  clustering_distance_colors= "manhattan",
  clustering_method= "average",
  border_color= NA,
  fontsize= 13
)

```

Use this heatmap
```{r}
# Make the annotation dataframe for the heatmap with 'Patient' column
heat_annot <- data.frame(
  row.names = targets$FileName,
  Gender = targets$genders,
  Treatment= targets$Condition,
  Patient = targets$Patient,  # Changed Treatment to Patient
  Tissue = targets$CleanCel
)

# Order the heat_annot dataframe by Gender
heat_annot <- heat_annot[order(heat_annot$Gender), ]
# Define colors for annotation color bars
patient_colors <- c("AW" = "#CEEAB9", "GF" = "#8EF9D0", "GH" = "#5BACF7", "MC" = "#E9B1F7", "NG" = "#B48BE2")

ann_colors <- list(
  Gender = c('FEMALE' = "#F21F66", 'MALE' = "#176EBD"), 
  Treatment = c('Baseline'= "grey", 'Post_Drug'= "black"),
  Patient = patient_colors,  # Use the patient_colors vector here
  Tissue = c('monocyte' = "#999BB2", 'neutrophil' = "#99B2A8", 'pbmc' = "#385CA2", 't.cell' = "#8A5851", 'whole.blood' = "#AE8B75")
)

# Extracting the top CpGs using topTable
top_cpgs <- topTable(fit2, number=Inf, coef="Post_Drug - Baseline", sort.by="p", p.value=0.05)

# Filtering CpGs based on effect size and select top 1000
selected_cpgs <- top_cpgs[abs(top_cpgs$logFC) > 0.05, ][1:1000,]

# Subset betas
selected_beta <- beta_subset[rownames(beta_subset) %in% rownames(selected_cpgs), ]

selected_beta_no_na <- selected_beta[complete.cases(selected_beta), ]
# Calculate variable of each CpG across all samples(all subjects, all tissue types)
library(matrixStats)
CpG_Var <- matrixStats::rowVars(selected_beta_no_na)
rankVar <- data.frame('order' = rank(-CpG_Var), 'var'= CpG_Var)
rankVar$top1000 <- ifelse(rankVar$order <= 1000, 'Yes', 'No')
ggpubr::ggscatter(rankVar, x= 'order', y= 'var', color= 'top1000', palette= c('black', 'red') )

# Get rank of CpGs by
rankvar= rank(-CpG_Var)

#colnames(selected_beta_no_na) <- targets$Patient[match(colnames(selected_beta_no_na), targets$Sample)]

# top 10,000 cpgs
data.topvar <- selected_beta_no_na[rankVar$top1000 == 'Yes',]
data.topvar <- as.matrix(data.topvar)
#(selected_beta_no_na) <- targets$Patient[match(colnames(selected_beta_no_na), targets$Sample)]



# Reorder the columns of data.topvar to match the order of heat_annot by Gender
#data.topvar <- data.topvar[, heat_annot$row.names]

#6 blocks of 8 if plotting by chip there might not be enough cpg's 
pheatmap(
  data.topvar,
  annotation_names_col= T,
  show_rownames= FALSE, #CpGs
  show_colnames= FALSE, #Samples
  annotation_col= heat_annot, 
  #annotation_row= row_annot,
  annotation_colors= ann_colors,
  cluster_cols= TRUE,
  color= colorRampPalette(c("yellow", "blue"))(1024),
  clustering_distance_rows= "manhattan",
  clustering_distance_colors= "manhattan",
  clustering_method= "average",
  border_color= NA,
  fontsize= 13
)
```

```{r}
# Order heat_annot correctly
heat_annot_ordered <- heat_annot[match(colnames(data.topvar), rownames(heat_annot)), ]

# Filter the data for baseline and post-drug treatments
baseline_data <- data.topvar[, heat_annot_ordered$Treatment == 'Baseline']
post_drug_data <- data.topvar[, heat_annot_ordered$Treatment == 'Post_Drug']

# Corresponding annotations for each subset
baseline_annot <- heat_annot_ordered[heat_annot_ordered$Treatment == 'Baseline', ]
post_drug_annot <- heat_annot_ordered[heat_annot_ordered$Treatment == 'Post_Drug', ]

# Check for NA, NaN, or Inf values and replace them if necessary
# For example, replacing NAs with the mean of non-NA values in the column
baseline_data[is.na(baseline_data)] <- apply(baseline_data, 2, function(x) mean(x, na.rm = TRUE))
post_drug_data[is.na(post_drug_data)] <- apply(post_drug_data, 2, function(x) mean(x, na.rm = TRUE))


# Check for NA, NaN, or Inf values in the baseline_data
sum(is.na(baseline_data))  # Should be 0 for no NAs
sum(is.nan(baseline_data)) # Should be 0 for no NaNs
sum(is.infinite(baseline_data)) # Should be 0 for no Infs


# Plot the baseline heatmap
#png("/Users/adiallo/Desktop/baseline_heatmap.png", width = 10, height = 10, units = "in", res = 300)
pheatmap(
  baseline_data,
  annotation_col = baseline_annot,
  annotation_colors = ann_colors,
  main = "Base Line",
  annotation_names_col= TRUE,
  show_rownames= FALSE,
  show_colnames= FALSE,
  cluster_cols= FALSE,  # You can cluster within the treatment group
  color = colorRampPalette(c("yellow", "blue"))(1024),
  border_color = NA,
  fontsize = 13
)
#dev.off()

# Plot the post-drug heatmap
#png("/Users/adiallo/Desktop/post_drug_heatmap.png", width = 10, height = 10, units = "in", res = 300)
pheatmap(
  post_drug_data,
  annotation_col = post_drug_annot,
  annotation_colors = ann_colors,
  main = "Post Drug",
  annotation_names_col= TRUE,
  show_rownames= FALSE,
  show_colnames= FALSE,
  cluster_cols= TRUE,  # You can cluster within the treatment group
  color = colorRampPalette(c("yellow", "blue"))(1024),
  border_color = NA,
  fontsize = 13
)
#dev.off()

```
```{r}
# Make the annotation dataframe for the heatmap with 'Patient' column
heat_annot <- data.frame(
  row.names = targets$FileName,
  Gender = targets$genders,
  Treatment= targets$Condition,
  Patient = targets$Patient,  # Changed Treatment to Patient
  Tissue = targets$CleanCel
)

# Order the heat_annot dataframe by Gender
heat_annot <- heat_annot[order(heat_annot$Treatment), ]
# Define colors for annotation color bars
patient_colors <- c("AW" = "#CEEAB9", "GF" = "#8EF9D0", "GH" = "#5BACF7", "MC" = "#E9B1F7", "NG" = "#B48BE2")

ann_colors <- list(
  Gender = c('FEMALE' = "#F21F66", 'MALE' = "#176EBD"), 
  Treatment = c('Baseline'= "grey", 'Post_Drug'= "black"),
  Patient = patient_colors,  # Use the patient_colors vector here
  Tissue = c('monocyte' = "#999BB2", 'neutrophil' = "#99B2A8", 'pbmc' = "#385CA2", 't.cell' = "#8A5851", 'whole.blood' = "#AE8B75")
)

# Extracting the top CpGs using topTable
top_cpgs <- topTable(fit2, number=Inf, coef="Post_Drug - Baseline", sort.by="p", p.value=0.05)

# Filtering CpGs based on effect size and select top 1000
selected_cpgs <- top_cpgs[abs(top_cpgs$logFC) > 0.05, ][1:10000,]

# Subset betas
selected_beta <- beta_subset[rownames(beta_subset) %in% rownames(selected_cpgs), ]

selected_beta_no_na <- selected_beta[complete.cases(selected_beta), ]
# Calculate variable of each CpG across all samples(all subjects, all tissue types)
library(matrixStats)
CpG_Var <- matrixStats::rowVars(selected_beta_no_na)
rankVar <- data.frame('order' = rank(-CpG_Var), 'var'= CpG_Var)
rankVar$top10000 <- ifelse(rankVar$order <= 10000, 'Yes', 'No')
ggpubr::ggscatter(rankVar, x= 'order', y= 'var', color= 'top10000', palette= c('black', 'red') )

# Get rank of CpGs by
rankvar= rank(-CpG_Var)

#colnames(selected_beta_no_na) <- targets$Patient[match(colnames(selected_beta_no_na), targets$Sample)]

# top 10,000 cpgs
data.topvar <- selected_beta_no_na[rankVar$top10000 == 'Yes',]
data.topvar <- as.matrix(data.topvar)
#(selected_beta_no_na) <- targets$Patient[match(colnames(selected_beta_no_na), targets$Sample)]



# First, sort your targets dataframe by gender
# This will also sort the Sample column in the same order
targets_ordered_by_treatment <- targets[order(targets$Condition,targets$Condition), ]

# Create an ordered vector of sample names based on the gender order
ordered_sample_names <- targets_ordered_by_treatment$FileName

# Now reorder the columns of your data matrix to match this order
# This assumes that your data matrix columns correspond to Sample names
ordered_data <- data.topvar[, ordered_sample_names]

#6 blocks of 8 if plotting by chip there might not be enough cpg's 
png("/Users/adiallo/Desktop/post_drug_heatmap.png", width = 10, height = 20, units = "in", res = 400)
pheatmap(
  ordered_data,
  annotation_names_col= T,
  show_rownames= FALSE, #CpGs
  show_colnames= FALSE, #Samples
  annotation_col = heat_annot[ordered_sample_names, ],
  #annotation_row= row_annot,
  annotation_colors= ann_colors,
  cluster_cols= FALSE,
  color= colorRampPalette(c("yellow", "blue"))(1024),
  clustering_distance_rows= "manhattan",
  clustering_distance_colors= "manhattan",
  clustering_method= "average",
  border_color= NA,
  fontsize= 13
)
dev.off()
```

```{r}


baseline_data <- selected_beta_no_na[, targets$Condition == "Baseline"]
treatment_data <- selected_beta_no_na[, targets$Condition == "Post_Drug"]

# Initialize an empty list to store the results
diff_list <- list()

# Loop over all samples in the 'beta_subset'
for (file_name in colnames(beta_subset)) {
  # Find the corresponding row in 'targets' for the current file_name
  target_row <- targets[targets$FileName == file_name, ]
  
  # Proceed only if the sample is 'On Treatment' (TimePoint 1 or 2)
  if (nrow(target_row) == 1 && target_row$TimePoint %in% c(1, 2)) {
    patient <- target_row$Patient
    cell_type <- target_row$CleanCell
    
    # Find the baseline file for the same patient and cell type (TimePoint 0)
    baseline_file <- targets$FileName[targets$Patient == patient & targets$CleanCell == cell_type & targets$TimePoint == 0]
    
    # Make sure there's exactly one baseline file
    if (length(baseline_file) == 1) {
      # Extract baseline values for the matched patient and cell type
      baseline_values <- beta_subset[, colnames(beta_subset) == baseline_file]
      # Extract treatment values for the current sample
      treatment_values <- beta_subset[, file_name]
      
      # Calculate the difference and store in diff_list
      diff_list[[file_name]] <- treatment_values - baseline_values
    }
  }
}

# Check the output
length(diff_list)  # Should be more than 0 if differences were calculated


```



```{r}

# Assuming targets and beta_subset are already defined and correct
# This step adds a 'Combination' column that includes the time point
#targets$Combination <- paste(targets$Patient, targets$CleanCell, targets$TimePoint, sep = "_")

# Initialize an empty list to store the results
diff_list <- list()
count = 0
# Loop over each patient and cell type combination
for (patient in unique(targets$Patient)) {
  for (cell_type in unique(targets$CleanCell)) {
    # Identify the baseline sample
    baseline_rows <- targets[targets$Patient == patient & targets$CleanCell == cell_type & targets$TimePoint == 0, ]
    
    # Proceed only if a baseline sample is found
    if (nrow(baseline_rows) == 1) {
      baseline_file <- baseline_rows$FileName
      
      # For each post-drug time point
      for (time_point in 1:2) {
        post_drug_rows <- targets[targets$Patient == patient & targets$CleanCell == cell_type & targets$TimePoint == time_point, ]
        
        # Loop through all post-drug samples for this time point
        for (j in 1:nrow(post_drug_rows)) {
          post_drug_row <- post_drug_rows[j, ]
          post_drug_file <- post_drug_row$FileName
          
          # Calculate the difference
          baseline_values <- selected_beta_no_na[, colnames(selected_beta_no_na) == baseline_file]
          treatment_values <- selected_beta_no_na[, colnames(selected_beta_no_na) == post_drug_file]
          diff <- treatment_values - baseline_values
          
          # Store the difference with a unique identifier
          combination_key <- paste(patient, cell_type, time_point, sep = "_")
          diff_list[[combination_key]] <- diff
          count = count + 1
        }
      }
    }
  }
}

cat("Total combinations processed:", count, "\n")
# Check the length of the diff_list to see how many combinations were processed
length(diff_list)


# Check the output
length(diff_list)  # Should be the number of differences calculated

# If you want to see the names of the list elements (which are the unique combinations)
names(diff_list)




```

```{r}
# Initialize an empty data frame for the long format
diff_data_long <- data.frame(
  Patient = character(),
  CellType = character(),
  TimePoint = integer(),
  Difference = numeric(),
  stringsAsFactors = FALSE
)

# Transform diff_list into diff_data_long
for (name in names(diff_list)) {
  parts <- strsplit(name, "_")[[1]]
  patient <- parts[1]
  cell_type <- parts[2]
  time_point <- as.numeric(parts[3])
  differences <- diff_list[[name]]
  v = cbind(cell_type,differences)

  if (length(differences) > 0) {
    temp_df <- data.frame(
      Patient = rep(patient, length(differences)),
      CellType = rep(cell_type, length(differences)),
      TimePoint = rep(time_point, length(differences)),
      Difference = differences,
      stringsAsFactors = FALSE
    )
   
    diff_data_long <- rbind(diff_data_long, temp_df)
  }
}

# Check the output for missing cell types
unique(diff_data_long$CellType)


```


```{r}
# Reset diff_list to ensure it's empty
diff_list <- list()

for (i in 1:nrow(targets)) {
  current <- targets[i, ]
  
  if (current$Condition == "Baseline") {
        baseline_index <- which(targets$Patient == current$Patient & 
                            targets$CleanCell == current$CleanCell & 
                            targets$TimePoint == 0)
        print(current$CleanCell)
        print(baseline_index)
  }
  # Proceed only if the sample is "On Treatment"
  if (current$Condition == "Post_Drug") {
    # Find the baseline sample for the same patient and cell type
    baseline_index <- which(targets$Patient == current$Patient & 
                            targets$CleanCell == current$CleanCell & 
                            targets$TimePoint == 0)
    #print(current$CleanCell)
    
    if (length(baseline_index) == 1) {
      print(paste("Baseline found for:", current$Patient, current$CleanCell))
    } else {
      print(paste("No baseline for:", current$Patient, current$CleanCell))
    }
    
    # Assuming you have a way to find post-drug samples, just as an example
    post_drug_index <- which(targets$Patient == current$Patient & 
                             targets$CleanCell == current$CleanCell & 
                             targets$TimePoint %in% c(1, 2))
                             
    if (length(post_drug_index) > 0) {
      for (idx in post_drug_index) {
        print(paste("Post-drug found for:", targets$Patient[idx], targets$CleanCell[idx], "TimePoint", targets$TimePoint[idx]))
      }
    }
  }
}


# Print diagnostic information
print(sprintf("Total combinations: %d", length(diff_list)))
print(names(diff_list))

# Specifically check for "NG" "monocyte"
ng_monocyte_combinations <- grep("NG_monocyte", names(diff_list), value = TRUE)
print(ng_monocyte_combinations)

ng_monocyte_baseline <- nrow(targets[targets$Patient == "NG" & targets$CleanCell == "monocyte" & targets$TimePoint == 0,])
ng_monocyte_postdrug <- nrow(targets[targets$Patient == "NG" & targets$CleanCell == "monocyte" & targets$TimePoint %in% c(1, 2),])

cat("NG monocyte Baseline samples:", ng_monocyte_baseline, "\n")
cat("NG monocyte Post-drug samples:", ng_monocyte_postdrug, "\n")


which(targets$Patient == current$Patient & 
                            targets$CleanCell == current$CleanCell & 
                            targets$Condition == "Baseline")
```

